AiPoker - An Implementation of a Poker Agent

### 2014-11-25 report/paper.tex
Correction report

\title{AiPoker - Implementation of a Poker Agent}

@@

%% indicate IEEE Member or Student Member in form indicated below
\author{John Holl\'en, Robin Berntsson, Simon Bergstr\"om}
\authorfooter{
%% insert punctuation at end of each item
\item
 John Holl\'en, Robin Berntsson, Simon Bergstr\"om
}

@@

\begin{document}
\pagestyle{fancy}
\rhead{TNM095}
\lhead{\empty}
\cfoot{\empty}
\rfoot{Page \ \thepage \ of \ \pageref{LastPage}}
\pagenumbering{arabic}
%% The ``\maketitle'' command must be the first command after the
%% ``\begin{document}'' command. It prepares and prints the title block.

@@

The classification in (table \ref{tab:table1}) shows that games like Chess and Go belong to the category of games that are less complex to solve with an AI agent. Poker belongs to one of the most complex games to solve, since it is both stochastic and contains imperfect information.

@@

\subsubsection{Expert Systems}
Billings et al.(1998)\cite{billingspapp} developed an AI agent called Loki. It is an expert system that uses a set of rules and formulas based on the strength of the cards to determine the next action. It uses no opponent modeling in heads up poker. It can be challenging to define strong expert systems as it can be hard to define rules that cover all aspects of the game. There can also be problems with conflicting rules and missing rules (situations that are not covered by any rules). Another problem is that expert systems require that you have great knowledge about the game as the AI will never play better than the defined the rules.

@@

\subsection{Monte Carlo Methods}
Monte Carlo methods is a class of algorithms that rely on random sampling to obtain a numerical estimate of a complicated problem. The method was invented by Stanislaw Ulam while working with the nuclear weapon program at Los Alamos in the 1940s \cite{Metropolis}. 
\begin{figure}[here]
  \begin{center}
    \includegraphics[scale=0.35]{img/integralexample.png}
    \caption{\label{fig:integralex} An illustration of how to determine the value of an integral with the help of Monte Carlo methods.}
  \end{center}
\end{figure}

One possible use for Monte Carlo methods is to determine the value of an integral as illustrated in fig(\ref{fig:integralex}). By randomly placing dots and label them as either under or over the function value an estimate of the area can be achieved. By taking the area of the square encapsulating the function times the ratio between the number of points under the function and the total amount of points. The approximated value given by a Monte Carlo method will converge to the correct value as the number of sample points goes to infinity\cite{Kleij}.

The insight that random sampling can be used to solve problems that are too complex to solve analytically, can be used to evaluate the best action in huge game trees. For a given node $N$, the value can be estimated by running a number of simulations of how the game may end from that node. The result from those simulations can then be used to give an approximation for the value of that node. As $N$ grows it will converge to the result from a Minimax search.

Monte Carlo tree search has made huge improvements for games that were previously very hard for AIs. One example is Go in which agents powered by MCTS now can compete with the best players in the world.

\section{Theory}
\subsection{Texas Hold'em no Limit}
\textit{(If you already are familiar with the game of Texas Hold'em, you can skip to section 2.2).}
There are many forms of poker, the most popular being Texas Hold'em. Texas Hold'em can be played in two different ways, limited and no-limit. In limited poker the bet and raise amount are fixed, there is also an upper limit of how many times the players raise. However in no-limit poker there is no limit for how much the players can bet, and there is also no limit for how many times the payers can raise.
\subsubsection{Main Concept}
The aim of the game is to win as much money as possible. When a player is out of money, that player has lost. When everyone except one player has run out of money the game is over and a winner is designated.

@@

\subsection{AI-techniques}
A couple of different AI-techniques were implemented in this study. They are described in this section. Also some tests where the different AIs compete against each other were made. The results from these tests are presented in the Results section.

\subsubsection{Monte Carlo Tree Search}
MCTS is a best-first search strategy. Every node $(N)$ in the tree contains two values, an estimate of the expected value $V(r(N))$ of the reward $r(N)$ and the number of times the node has been visited $n_{i}$. MCTS only starts with the root node and incrementally builds the tree by repeating the following 4 steps:\\*\\*
\textbf{Selection}
\\*Selection is made by starting from the root node and then recursively select nodes until a leaf node is reached (this does not have to be a leaf of the game tree). To determine which node is the optimal one, the upper confidence bounds formula (eq(\ref{eq:upperconfidence})) is used. \cite{levente}:

\begin{equation} \label{eq:upperconfidence}
	v_{i} + C \cdot \sqrt{\frac{\ln N}{n_{i}}}
\end{equation}
The first term $v_{i}$ is the estimated value of the node and is responsible for exploitation of good nodes, the second term is made up by the total number of visits of the parent node $N$ and the total number of visits of the current node $n_{i}$. The constant $C$ allows tuning of the exploration-exploitation trade-off. 
\\*\\*
\textbf{Expansion}
\\*If the leaf node reached is not a terminating node, which means that the game is not over at the current node. Then create one or more child nodes from the node currently at. This step is illustrated in fig(\ref{fig:expansion}).
\begin{figure}[here]
  \begin{center}
    \includegraphics[scale=0.28]{img/tree1.png}
    \caption{\label{fig:expansion} Illustration of how the green node is added to the tree.}
  \end{center}
\end{figure}
\\*\\*
\textbf{Simulation}
\\*The third step in MCTS is the simulation step. Here a game is simulated from the state in the new node the game has ended. The value of the reached result (won amount of money) is recorded.
\\*\\*
\textbf{Backpropagation}
\\*The last step is to update the expected value and the selection counter for all nodes along the explored path by backpropagating the recorded result from the simulation step. This step is illustrated in fig(\ref{fig:backprop}).
\begin{figure}[here]
  \begin{center}
    \includegraphics[scale=0.28]{img/tree2.png}
    \caption{\label{fig:backprop} Illustration of how the path to the node is updated with the score from the simulation in the whole tree.}
  \end{center}
\end{figure}
When the back propagation has been done the whole chain starts over again starting with the selection stage\cite{Chaslot}. The use of the upper bounds confidence selection method makes the game tree grow asymmetrically as it focus its search in the more promising parts of the tree.

@@

\subsubsection{Bucketing}
Since effective hand strength (EHS) needs at least the flop to be revealed in order to be used for decision making, the Bucketing algorithm was implemented  as a complement to the EHS algorithm. This algorithm is only used in in the pre-flop phase where no cards are visible on the table \cite{schweizer}.
\begin{figure}[here]
  \begin{center}
    \includegraphics[scale=0.60]{img/bucketlist.png}
    \caption{\label{fig:bucketlist} The bucket list.}
  \end{center}
\end{figure}
\\*The table in fig(\ref{fig:bucketlist}) gives every two card combination a value between 0 and 4 where 4 is a good combination and 0 is a bad combination. This list is the source of decision making in the pre-flop phase when using the EHS-algorithm as logic in the poker agent.

@@

The application is implemented in an object oriented manner. The main application file is called main.js and it is here the scene is created and where all the other classes needed are initialized. The file main.js also contains all the click listeners for the buttons the human player uses to do moves. An overview over the game architecture is shown in fig(\ref{fig:arcitect}). 
\begin{figure}[here]
  \begin{center}
    \includegraphics[scale=0.50]{img/arcitect.png}
    \caption{\label{fig:arcitect} A brief overview of the application architecture.}
  \end{center}
\end{figure}

@@

\section{Result}
\subsection{The Game}
The result of this study is a Texas Hold'em game with an AI implemented using several different AI techniques combined together. The AI in the interactive game is using Monte Carlo Tree Search together with Hand Strength as a heuristic. The game is fully playable with an opponent that is realistic in the way it chooses what move to make. The 3D-scene is used to visualize the poker table, and it also shows the cards on hand. This is shown in fig(\ref{fig:3dscene}).
\begin{figure}[H]
  \begin{center}
    \includegraphics[scale=0.27]{img/3dscene.png}
    \caption{\label{fig:3dscene} The 3D-scene in the final version showing the player's cards on hand.}
  \end{center}
\end{figure}

To the left of the 3D-scene there are some additional UI-elements. these are the buttons telling the player what moves he or she can make. By pressing one of these buttons the player will make a move. The coloured buttons indicate that a move is valid. The greyed out buttons indicate invalid moves. The buttons are shown in fig(\ref{fig:ui1}).
\begin{figure}[H]
  \begin{center}
    \includegraphics[scale=0.50]{img/ui1.png}
    \caption{\label{fig:ui1} The player controls, placed to the left of the 3D-scene. Greyed out buttons indicate an invalid move at the current state.}
  \end{center}
\end{figure}

To the right of the 3D-scene some information is shown. Information about money is visible in the upper box. How much money each player has, and how much money there is currently on the table. This information box is illustrated in fig(\ref{fig:ui2}).
\begin{figure}[H]
  \begin{center}
    \includegraphics[scale=0.50]{img/ui2.png}
    \caption{\label{fig:ui2} The score board giving the player information about the money currently in play.}
  \end{center}
\end{figure}

>>>>

\title{AiPoker - An Implementation of a Poker Agent}

@@

%% indicate IEEE Member or Student Member in form indicated below
\author{John Holl\'en, Robin Berntsson, Simon Bergstr\"om}
\authorfooter{
%% insert punctuation at end of each item
\item
 Simon Bergstr\"om, Media Technology student at Link\"oping University, Sweden, e-mail: simbe109@student.liu.se.
\item
  Robin Berntsson, Media Technology student at Link\"oping University, Sweden, e-mail: robbe482@student.liu.se.
\item
  John Holl\'en, Media Technology student at Link\"oping University, Sweden, e-mail: johho982@student.liu.se.
}

@@

\begin{document}

\fancypagestyle{empty}{
 \fancyhf{}
 \fancyhead[R]{TNM095 \\* HT1 2014}
 \fancyhead[C]{Artificial Intelligence for Interactive Media}
}

\pagestyle{fancy}
\rhead{TNM095 \\* HT1 2014}
\chead{Artificial Intelligence for Interactive Media}
\lhead{\empty}
\cfoot{\empty}
\rfoot{Page \ \thepage \ of \ \pageref{LastPage}}
\pagenumbering{arabic}
%% The ``\maketitle'' command must be the first command after the
%% ``\begin{document}'' command. It prepares and prints the title block.

@@

The classification in Table.\ref{tab:table1} shows that games like Chess and Go belong to the category of games that are less complex to solve with an AI agent. Poker belongs to one of the most complex games to solve, since it is both stochastic and has imperfect information.

@@

\subsubsection{Expert Systems}
Billings et al.(1998)\cite{billingspapp} developed an AI agent called Loki. It is an expert system that uses a set of rules and formulas based on the strength of the cards to determine the next action. It uses no opponent modeling in two player poker. It can be challenging to define strong expert systems as it can be hard to define rules that cover all aspects of the game. There can also be problems with conflicting rules and missing rules (situations that are not covered by any rules). Another problem is that expert systems require that you have great knowledge about the game as the AI will never play better than the defined the rules.

@@

\subsection{Monte Carlo Methods}
Monte Carlo methods is a class of algorithms that rely on random sampling to obtain a numerical estimate of a complex problem. The method was invented by Stanislaw Ulam while working with the nuclear weapon program at Los Alamos in the 1940s \cite{Metropolis}. 
\begin{figure}[here]
  \begin{center}
    \includegraphics[scale=0.35]{img/integralexample.png}
    \caption{\label{fig:integralex} An illustration of how to determine the value of an integral with the help of Monte Carlo methods.}
  \end{center}
\end{figure}

One possible use for Monte Carlo methods is to determine the value of an integral as illustrated in Fig.\ref{fig:integralex}. By randomly placing dots and labelling them as either under or over the function value an estimate of the area can be achieved. By taking the area of the square encapsulating the function times the ratio between the number of points under the function and the total amount of points. The approximated value given by a Monte Carlo method will converge to the correct value as the number of sample points goes to infinity\cite{Kleij}.

The insight that random sampling can be used to solve problems that are too complex to solve analytically can be used to evaluate the best action in huge game trees. For a given node $N$, the value can be estimated by running a number of simulations of how the game may end from that node. The result from those simulations can then be used to give an approximation for the value of that node. As the number of nodes grows it will converge to the result from a Minimax search.

Monte Carlo tree search (MCTS) has made huge improvements for games that were previously very hard for AIs. One example is Go in which agents powered by MCTS now can compete with the best players in the world.

\section{Theory}
\textit{(If you already are familiar with the game of Texas Hold'em, you can skip to section 2.2).}
\subsection{Texas Hold'em no Limit}
There are many forms of poker, the most popular being Texas Hold'em. Texas Hold'em can be played in two different ways, limited and no-limit. In limited poker the bet and raise amount are fixed, there is also an upper limit of how many times the players raise. However in no-limit poker there is no limit for how much the players can bet, and there is also no limit for how many times the payers can raise.
\subsubsection{Main Concept}
The aim of the game is to win as much money as possible. When a player is out of money, that player has lost. When everyone except one player has run out of money the game is over and a winner is designated.

@@

\subsection{AI-techniques}
A couple of different AI-techniques were implemented in this study. They are described in this section. Also some tests where the different AIs compete against each other were made. The results from these tests are presented in the Result section.

\subsubsection{Monte Carlo Tree Search}
MCTS is a best-first search strategy. Every node $N$ in the tree contains two values, an estimate of the expected reward $v_{i}$ and the number of times the node has been visited $n_{i}$. MCTS only starts with the root node and incrementally builds the tree by repeating the following 4 steps:\\*\\*
\textbf{Selection}
\\*Selection is made by starting from the root node and then recursively select nodes until a leaf node is reached (this does not have to be a leaf of the game tree). To determine which node is the optimal one, the upper confidence bounds formula Eq.\ref{eq:upperconfidence} is used \cite{levente}.

\begin{equation} \label{eq:upperconfidence}
	v_{i} + C \cdot \sqrt{\frac{\ln (N)}{n_{i}}}
\end{equation}
The first term $v_{i}$ is the estimated reward of the node and is responsible for exploitation of good nodes, the second term is made up by the total number of visits of the parent node $N$ and the total number of visits of the current node $n_{i}$. The constant $C$ allows tuning of the exploration-exploitation trade-off. 
\\*\\*
\textbf{Expansion}
\\*If the leaf node reached is not a terminating node, which means that the game is not over at the current node. Then create one or more child nodes from the node currently at. This step is illustrated in Fig.\ref{fig:expansion}.
\begin{figure}[here]
  \begin{center}
    \includegraphics[scale=0.28]{img/tree1.png}
    \caption{\label{fig:expansion} Illustration of how the green node is added to the tree.}
  \end{center}
\end{figure}
\\*\\*
\textbf{Simulation}
\\*The third step in MCTS is the simulation step. Here a game is simulated from the state in the new node until the game has ended. The value of the reached result (won amount of money) is recorded.
\\*\\*
\textbf{Backpropagation}
\\*The last step is to update the expected value and the number of times visited for all nodes along the explored path by backpropagating the recorded result from the simulation step. This step is illustrated in Fig.\ref{fig:backprop}.
\begin{figure}[here]
  \begin{center}
    \includegraphics[scale=0.28]{img/tree2.png}
    \caption{\label{fig:backprop} Illustration of how the path to the node is updated with the score from the simulation in the whole tree.}
  \end{center}
\end{figure}
When the back propagation has been done the whole chain starts over again starting with the selection stage\cite{Chaslot}. The use of the upper bounds confidence selection method makes the game tree grow asymmetrically as it focuses its search in the more promising parts of the tree.

@@

\subsubsection{Bucketing}
Since effective hand strength (EHS) needs at least the flop to be revealed in order to be used for decision making, the Bucketing algorithm was implemented  as a complement to the EHS algorithm. This algorithm is only used in in the pre-flop phase where no cards are visible on the table \cite{schweizer}.
\begin{figure}[here]
  \begin{center}
    \includegraphics[scale=0.60]{img/bucketlist.png}
    \caption{\label{fig:bucketlist} The bucket list.}
  \end{center}
\end{figure}
\\*The table in Fig.\ref{fig:bucketlist} gives every two card combination a value between 0 and 4 where 4 is a good combination and 0 is a bad combination. This list is the source of decision making in the pre-flop phase when using the EHS-algorithm as logic in the poker agent.

@@

The application is implemented in an object oriented manner. The main application file is called main.js and it is here the scene is created and where all the other classes needed are initialized. The file main.js also contains all the click listeners for the buttons the human player uses to do moves. An overview over the game architecture is shown in Fig.\ref{fig:arcitect}. 
\begin{figure}[here]
  \begin{center}
    \includegraphics[scale=0.50]{img/arcitect.png}
    \caption{\label{fig:arcitect} A brief overview of the application architecture.}
  \end{center}
\end{figure}

@@

\section{Result}
\subsection{The Game}
The result of this study is a Texas Hold'em game with an AI implemented using several different AI techniques combined together. The AI in the interactive game is using Monte Carlo Tree Search together with Hand Strength as a heuristic. The game is fully playable with an opponent that is realistic in the way it chooses which move to make. The 3D-scene is used to visualize the poker table, and it also shows the cards on hand. This is shown in Fig.\ref{fig:3dscene}.
\begin{figure}[H]
  \begin{center}
    \includegraphics[scale=0.27]{img/3dscene.png}
    \caption{\label{fig:3dscene} The 3D-scene in the final version showing the player's cards on hand.}
  \end{center}
\end{figure}

To the left of the 3D-scene there are some additional UI-elements. these are the buttons telling the player what moves he or she can make. By pressing one of these buttons the player will make a move. The coloured buttons indicate that a move is valid. The greyed out buttons indicate invalid moves. The buttons are shown in Fig.\ref{fig:ui1}.
\begin{figure}[H]
  \begin{center}
    \includegraphics[scale=0.50]{img/ui1.png}
    \caption{\label{fig:ui1} The player controls, placed to the left of the 3D-scene. Greyed out buttons indicate an invalid move at the current state.}
  \end{center}
\end{figure}

To the right of the 3D-scene some information is shown. Information about money is visible in the upper box. How much money each player has, and how much money there is currently on the table. This information box is illustrated in Fig.\ref{fig:ui2}.
\begin{figure}[H]
  \begin{center}
    \includegraphics[scale=0.50]{img/ui2.png}
    \caption{\label{fig:ui2} The score board giving the player information about the money currently in play.}
  \end{center}
\end{figure}
